\chapter*{Acknowledgements}


First, I would like to thank my supervisors Prof. José F. Aldana Montes and Prof. Antonio J. Nebro Urbaneja for their guidance and support throughout these years of developing the work that is part of this thesis. I would especially like to thank them for giving me the opportunity of doing a PhD thesis when I didn't think it was possible for me to do it.

I also wish to thank everyone who accepted to be part of my thesis committee, for agreeing so quickly, and making it all so easy for me. In addition I thank my external evaluators for all their insightful corrections that improved my manuscript.

My sincere thanks to Dr. Manuel López Ibañez, Dr. Julia Handl and Dr. Richard Allmendinger for the lovely stay I had in those three months at the University of Manchester, in particular Manuel who I worked with and who helped me in my research.

I would like to mention all the people I have been working with all these years in the Khaos research group, both doctors and students. Most of them I am glad to call friends, as we have shared lots of fun and hard work over the years.

I would also like to thank other members of the Grupo de Ingeniería del Software de la Universidad de Málaga (GISUM) who I know personally and have shared personal experiences with them. I want to thank Lisa Huckfield for all her corrections in all my papers, doing them faster than I thought humanly possible. In addition, I want to thank all the staff at the Ada Byron research centre for our little chats that we had after leaving late from work.

I don't want to forget to thank all my family, especially my parents, my brother, my two little sisters, and even my cat, thanks for always being there and making me smile in difficult times.

And last but not least, I would like to thank above all my girlfriend (soon to be Dr also) María Jesús García Godoy. The most hard working person I have ever known and the only one I know who can achieve whatever she puts her mind to. Without her support (both personal and professional) I would never have been able to complete this PhD thesis.
